\documentclass[../full_thesis/full_thesis.tex]{subfiles}

% Default image directory
\newcommand{\thisdir}{../introduction}
\graphicspath{{\thisdir/img/}} 

\begin{document} 

Neutron stars were first postulated by Landau as `dense stars
which look like giant atomic nuclei' \citep{Yakovlev2013}, even before the
discovery of the neutron by \citet{Chadwick1932}. However it was
\citet{Baade1934} who made the explicit prediction of a neutron star whilst
trying to explain the energy released in observed supernova.

In a main-sequence star, the nuclear fusion of
hydrogen atoms into helium provides outward pressure balancing the star in an
equilibrium configuration with the inward pressure of the stars self-gravity.
Eventually the star depletes its reserves of hydrogen and can no longer
maintain equilibrium. If the star has an initial mass greater than $\sim 8
\Msun$, then it may undergo a \emph{core-collapse supernova} during which some
of the mass is ejected, but the rest falls in creating a new compact object.
In this object, temperatures and pressure rapidly sore and the
electrons and protons undergo inverse beta decay combining to form neutrons and
neutrinos:
\begin{equation}
    e^{-} + p \rightarrow n + \nu.
\end{equation}
If the compact objects mass is less than $\sim 5 \Msun$ then, once the pressures
reach nuclear densities of $\sim 2.3 \times10^{14}$~g/cm$^{3}$, neutron
degeneracy pressure can halt the collapse in a new compact stable equilibrium
configuration which we call a neutron star. For remnants with larger masses,
this is not possible and the object will collapse to form a black-hole; the
detail of exactly what the critical mass a neutron star can sustain is
sensitive to the equation of state of matter under these conditions.

Our knowledge of neutron stars is founded on observations made by
electromagnetic astronomy. This has revealed a wealth of different neutron
stars and their phenomena which we will introduce in the following sections.
Many of these observed phenomena have well defined models which allow us to
infer properties of neutron stars themselves.
However, our knowledge of neutron stars is far from complete: current
observations can be contradictory or have features not explained by any known
physical models. Improvements in electromagnetic astronomy will bring to light
a greater number of new neutron stars and improve the resolution of those
currently observed; it is hoped that this will help us to better understand them.

There are two other methods we can utilise to learn more about neutron stars: improved
modelling of current observations and by observing them from their
gravitational wave emissions. In this thesis, we will study how one of the
observed phenomena, so-called `timing variations', can help us to learn more
from current observations and also test whether it may hinder the current
search for gravitational waves from neutron stars.

In this introduction, we will acquaint the reader with the current observations
of neutron star and introduce some of the basic physics.

\section{Observation of pulsars and their identification with neutron stars}
\input{\thisdir/observation_of_pulsars}

\section{Pulsar timing}
\label{sec: pulsar timing methods}
\input{\thisdir/pulsar_timing_methods}

\section{Categorising neutron stars}
\label{sec: categorising neutron stars}
\input{\thisdir/categorising_neutron_stars}

\section{Radio pulsar population statistics}
\label{sec: population stats}
\input{\thisdir/population_stats}

\section{The physics of rotation powered pulsars} 
\label{sec: rotation powered pulsars}
\input{\thisdir/rotation_powered_pulsars}

\section{Neutron stars and gravitational waves}
\label{sec: gravitational waves}
\input{\thisdir/gravitational_waves}


\section{Plan of the thesis}

Following this introductory chapter, we will have a chapter introducing
so-called timing variations in pulsars: glitches and timing-noise. This will
introduce the observed phenomena and describe the current state of modelling.
We will provide some original work on simple ways which the models could be
tested.

In the next four chapters we evaluate models of timing noise in the face of
current observations and attempt to constrain the models.  In Chapter~\ref{sec:
rotating frame} we explore how \emph{precession}, a potential ingredient to
explain timing-noise, operates in the inertial frame of the star.  Following
this, Chapter~\ref{sec: rotating frame} looks at how precession will manifest
in the observations made by pulsar astronomers. In Chapter~\ref{sec: testing
models} we perform a rigorous quantitative model comparison between precession and the
leading alternative, \emph{switching}, for PSR~B1828-11. This is then extended
in Chapter.~\ref{sec: extending precession models} where we use the model
comparison tools to discover previously unknown fundamental physics of the
pulsar.

In the final three chapters we approach another important aspect of timing
variations for neutron stars: the effect they may have on our ability to
detection gravitational waves from neutron stars. In Chapter~\ref{sec: intro to
cgw} we introduce the methods and formalisms used by gravitational wave
astronomy before analysing the effect of glitches in Chapter~\ref{sec: glitches
in cgw} and the effect of timing-noise in Chapter~\ref{sec: timing noise in
cgw}.

\biblio


\end{document}
