\documentclass[../full_thesis/full_thesis.tex]{subfiles}

% Default image directory
\newcommand{\thisdir}{../analytic_timing_noise_cgw}
\graphicspath{{\thisdir/img/}}

\begin{document}

\meta{
It's unclear at this time what causes timing noise, but from Chapter.~\ref{sec:
timing noise in cgw} we know that it may pose a problem for continuous GW
searches if the phase evolution is effected by timing noise and that is not
included in the search templates. In addition to searches for isolated neutron
stars, many searches \citep[see for
example][]{ligo2015scox1,leaci2015,ScoX1:MDC1} have been performed for GWs from
low-mass X-ray binary systems (LMXBs). These searches experience similar
difficulties due to a stronger form of timing noise known as `spin-wandering'.
This issue was investigated by \citet{watts2008} who quantifed the effect using
a decoherence time $T_{\textrm{decoh}}$, after which the Taylor expansion can
no longer track the phase, defined by
\begin{align}
T_{\textrm{decoh}}^{2} \dot{\nu}_{\textrm{s}} = 1.
\end{align}
They then estimated a worst-case decoherence time by using the maximimal
spin-up rate due to the accretion torque and found that for some sources such
as Sco-X1 the decoherence time can be short as $\sim 1$~week.

In this chapter, we will present some preliminary calculations and results
related to modelling timing noise in a continuous GW using a random walk model.
Unlike the results of Chapter.~\ref{sec: timing noise in cgw}, which used an
empirical description of timing noise given by the Crab ephemeris, the results
derived here can be applied to any search in which it is thought the signal may
undergo a random walk. In this sense, it is equivalent to Chapter~\ref{sec:
glitches in cgw} in that the ultimate aim is to use observations to develop
fitting formulae for the random walk model and then predict the amount of
mismatch, given the search parameters. As given here, this task in incomplete
and so we will present our initial findings with a view to develop this chapter
further in the future.

Recent observations by \citet{Hobbs2010} suggest that a random walk model does
not capture the physics of timing noise in isolated radio pulsars and hence is
not a useful way to infer neutron star physics. Nevertheless, the random walk
model remains a practical empirical model; in this section then, we use it as
such without requiring it to have any deeper substantive meaning for what
causes timing noise. In the same way, the random walk model can also be applied
to LMXBs where the amount of spin-wandering could be inferred from fluctuations
in the luminosity. From this we intend in the future to update the estimates by
\citet{watts2008} to estimate the mismatch for various searches.

}

\section{Defining a random walk}
\label{sec: Defining a random walk}
\input{\thisdir/defining_a_random_walk}

\section{Random walk models: a simple treatment}
\label{sec: random walk models part I}
\input{\thisdir/random_walk_models_part_I}

\section{Random walk models: minimising the mismatch} 
\label{sec: random walk models part II}
\input{\thisdir/random_walk_models_part_II}

\section{Conclusion}

In this section, we have calculated the expectation of the fully-coherent
mismatch when searching for a GW signal which undergoes a random walk in one of
the phase, frequency, or spin-down rate. We did this initially for an
unminimised system and then minimised with respect to the template parameters
$f_\textrm{t}$ and $\dot{f}_\textrm{t}$. In the future, we would like to develop
this work to make direct predictions for GW searches. For example,
searches for GWs from LMXBs assume a constant frequency, but torque fluctuations
could produce a random walk in the frequency; we could model this and, given
an estimate for the random walk parameters, estimate for what observation span
GW searches may be at risk of loosing the signal.

Another observation which we would like to understand further is the
implication of the power-law dependence on the observation time. In particular,
in Eqn.~\eqref{eqn: Expected mismatch RW in frequency k2} we found that for
a random walk in the frequency
$E[\mutilde] \propto \Tobs ^{3}$ while in Chapter.~\ref{sec: timing noise in
cgw} we found that the expected fully-coherent mismatch for timing noise as
described by the Crab ephemeris scales as $\propto \Tobs^{2.88\pm-0.03}$ (see
Eqn.~\eqref{eqn: fit values}). This suggests that the data in Crab ephemeris
may be consistent with a random walk in the frequency (something that was
first noted by \citet{Boynton1972}), but requires further investigation to
establish this.


\begin{subappendices}
\input{\thisdir/summation_identities}

\section{Least-squares minimisation of a random walk}
\label{sec: least squares minimisation of a random walk}
\input{\thisdir/FittingToARandomWalk_substance}


\end{subappendices}


\biblio

\end{document}
